---
title: "Exploratory Data Analysis"
output: 
  rmarkdown::github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Introduction

This section of the project will focus on exploratory data analysis. After examining and performing any necessary data cleaning, I visualized the variables and their relationships to determine whether or not they will be useful in predicting customer churning.

**Key Findings**:

- After data cleaning, the finalized training set contains 1995 observations with 20 variables
- The response variable, **churn**, has a prevalence of 27.3%
- Customers who have a higher monthly charge and lower tenure tend to churn more.
- Customers who are not senior citizens, have partners, have dependents, or don't use electronic checks are less likely to churn.
- Although customers with the *Fiber Optic* internet service are more likely to churn, it's due to the higher cost of that specific service rather than the quality of its services.
- Customers with online security, online backup, device protection, and tech support are less likely to churn regardless of the cost and their internet service.
- Based on these results, I decided to include these variables for predictive modeling.

### 1.1 Loading Packages

These are the packages I used during when exploring and visualizing the data.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Load required libraries
library(ggplot2)
library(gridExtra)
library(plyr)
library(dplyr)
library(GGally)

# Set seed for reproducibility
set.seed(101)
```

### 1.2 Importing the Data

The data has already been split into a training and testing set. I will be using the training set for exploratory data analysis and model building. The testing set will be reserved for determining model performance, but I will be verifying below that it has the same structure as the training set.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Import data
load("data/train.Rdata")
load("data/test.Rdata")
```

## 2. Data Cleaning

Prior to exploring the dataset, I need to examine the data and perform any necessary cleaning.

### 2.1 Viewing the Data

I viewed the structure of the data below and made the following observations:

- There are 2000 observations and 21 variables in each set of data, with **churning** being the response variable
- There are only 3 continuous variables: **tenure**, **monthly charges**, and **total charges**; the rest are factor variables
- Some factor variables depend on the value of another variable. For example, **online security** has three levels, "Yes", "No", and "No internet service", whose value depends on **internet service**
- **Customer ID** appears to be a unique identifier

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Preview the data
head(train, 3)

# View the structure of the training set
str(train)

# Verify that the structure of the testing set matches the training set
str(test)
```

### 2.2 Missing & Duplicate Values

Looking at the missing values, the training set contains 5 missing values for **total charges**. Upon further inspection of the data, the customers with these missing values have a tenure of 0, which means they had just began their subscription service. Thus, they will not be useful in predicting churning and will be dropped from the data set.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Check the rows with missing values
train[!complete.cases(train), ]

# Drop the missing values
train <- train[complete.cases(train), ]
```

I also checked for any potential duplicate observations and found that there are no duplicate values in the dataset.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Check for duplicated rows
dim(train[duplicated(train),])[1]
```

### 2.3 Data Preparation

Finally, I dropped the **customer ID** variable since it is a unique identifier that will not be useful for prediction. I also refactored **senior citizenship** to match the levels of the other binary variables for consistency in visualizations.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Drop the CustomerID variables
train <- subset(train, select = -c(customerID))

# Change SeniorCitizen to match the levels of the other binary variables
train$SeniorCitizen <- as.factor(mapvalues(train$SeniorCitizen, from=c("0","1"), to=c("No", "Yes")))
```

### 2.4 Finalized Dataset

I looked at the structure of the data one last time to verify the changes I've made.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# View structure of the finalized dataset
str(train)
```

I also looked at the numerical summary of the variables in the dataset and found that 23.7% of customers ended up churning from the company.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# View a statistical summary of the finalized dataset
summary(train)

# Calculate prevalence of churning
544 / (1451 + 544)
```


## 3. Analyzing Relationships with Churning

Now that I have familiarized myself with the structure of the data and performed any necessary data cleaning, I can explore and visualize the data.

### 3.1 Distributions of Continuous Variables

The first set of variables I'm going to explore are the continuous variables. I plotted box plots of **tenure**, **monthly charges**, and **total charges** to compare their distributions between customers who churned and those who did not.

```{r, echo = TRUE, warning = FALSE, message=FALSE}
# Define function to plot a box plot
plot.boxplot <- function(data, x, y, x_label, y_label) {
ggplot(data, aes(x = x, y = y)) + 
  geom_boxplot(aes(fill = x)) + 
  xlab(x_label) + 
  ylab(y_label) +
  theme(axis.text=element_text(size=12), axis.title=element_text(size=12),         legend.position="none")
}
```

```{r, echo = TRUE, warning = FALSE, message=FALSE}
# Plot box plots of the continuous variables
bp1 <- plot.boxplot(train, train$Churn, train$tenure, "Churn", "Tenure")
bp2 <- plot.boxplot(train, train$Churn, train$MonthlyCharges, "Churn", "Monthly Charges")
bp3 <- plot.boxplot(train, train$Churn, train$TotalCharges, "Churn", "Total Charges")
grid.arrange(bp1, bp2, bp3, ncol=3)
```


**Findings:**

- Customers with a high **tenure** are less likely to churn
- Customers with a high **monthly charge** are more likely to churn
- The distributions of these variables for churning customers tend to be strongly skewed.

While customers who churn tend to have a lower total charge, this may be due to them having less time to accumulate a higher total charge. 

Next, I looked at the numerical summary of the continuous variables for each group of customers.

```{r, echo = TRUE, warning = FALSE, message=FALSE}
# Subset the data to include only the numeric variables
train.continuous <- subset(train, select=c(tenure, MonthlyCharges, TotalCharges, Churn))

# Calculate summary statistics of numeric variables for customers who churned
print(summary(train.continuous[train.continuous$Churn == 'Yes', ]))

# Calculate summary statistics of numeric variables for customers who did not churn
print(summary(train.continuous[train.continuous$Churn == 'No', ]))
```

Customers who churn have a median monthly charge of $79.60 dollars and a median tenure of 18 months. Customers who did not churn have a median monthly charge of \$64.05 and a median tenure of 37 months. As expected, customers who stay with the company will accumulate a longer tenure. In addition, they tend to be satisfied with the amount they are paying every month.

### 3.2 Bivariate Analysis of Continuous Variables

Next, I wanted to see how **tenure**, **monthly charges**, and **total charges** are related to each other. I plotted a scatterplot matrix of these variables to determine their relationships.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Plot scatterplot matrix of continuous variables
ggpairs(train.continuous, columns=1:3, ggplot2::aes(color=Churn, alpha=0.5))
```

The following relationships are shown in the scatterplot matrix:

- **Total charges** has a strong, positive relationship with **tenure** (*r = 0.827*)
- **Total charges** has a moderate, positive relationship with **monthly charges** (*r = 0.648*)
- **Monthly charges** has a weak, positive relationship with **tenure** (*r = 0.243*)

The scatterplots with **total charges** depict a triangular shape where no points lie above the diagonal line, *y = x*. This indicates a changing variance (heteroscedasticity), possibly due to another factor variable. These variables may also have the following relationship: *monthly charges* * *tenure* ≅ *total charges*.

To verify this relationship, I created a new variable, **charge difference**, which is equal to *total charges* - (*monthly charges* * *tenure*). I plotted its distribution below.

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="75%"}
# Create charge difference variable
train$ChargeDifference = train$TotalCharges - (train$MonthlyCharges * train$tenure)

# Plot distribution of charge difference
ggplot(train, aes(x=ChargeDifference)) + 
  geom_histogram(color="black", fill="white") + 
  geom_vline(aes(xintercept=mean(ChargeDifference)), color="blue", linetype="dashed", size=1) +
  xlab("Total Charges - (Monthly Charges * Tenure)") +
  ylab("Count") +
  ggtitle("Histogram of Charge Difference")
```

As expected, the result is symmetric and unimodal with a peak at 0, verifying the relationship. To avoid issues with collinearity, I will only include **tenure** and **monthly charges** in the model.

### 3.3 Analyzing Factor Variables

I plotted the remaining variables against churning.

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="50%"}
# Define function to plot bar plot
plot.barplot <- function(data, x, y, x_label, y_label, angle) {
  ggplot(data, aes(x = x)) + 
  geom_bar(aes(fill = y), position = "fill") + 
  xlab(x_label) +
  ylab(y_label) + 
  scale_fill_discrete(name = "Churn") +
  theme(axis.text=element_text(size=9), axis.title=element_text(size=9), axis.text.x = element_text(angle = angle))
}
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Plot bar plots of binary variables against churning
bp1 <- plot.barplot(train, train$gender, train$Churn, "Gender", "Proportion", 0)
bp2 <- plot.barplot(train, train$SeniorCitizen, train$Churn, "Senior Citizen", "Proportion", 0)
bp3 <- plot.barplot(train, train$Partner, train$Churn, "Partner", "Proportion", 0)
bp4 <- plot.barplot(train, train$Dependents, train$Churn, "Dependents", "Proportion", 0)
bp5 <- plot.barplot(train, train$PhoneService, train$Churn, "Phone Service", "Proportion", 0)
bp6 <- plot.barplot(train, train$PaperlessBilling, train$Churn, "Paperless Billing", "Proportion", 0)
grid.arrange(bp1, bp2, bp3, bp4, bp5, bp6, ncol=3)


```

There appears to be a significant difference in the proportion of customers who churned for the variables **senior citizen**, **partner**, **dependents**, and **paperless billing**. However, the proportion of customers who churned appear to be the same for **gender** and **phone service**.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Plot bar plots of factor variables against churning
bp7 <- plot.barplot(train, train$InternetService, train$Churn, "Internet Service", "Proportion", 15)
bp8 <- plot.barplot(train, train$Contract, train$Churn, "Contract", "Proportion", 15)
bp9 <- plot.barplot(train, train$MultipleLines, train$Churn, "MultipleLines", "Proportion", 15)
bp10 <- plot.barplot(train, train$PaymentMethod, train$Churn, "Payment Method", "Proportion", 15)
grid.arrange(bp7, bp8, bp9, bp10, ncol=2)
```

Customers who tend to churn the most have:

- *Fiber Optic* as their internet service
- a *month-to-month* contract
- *electronic check* as their payment method.


```{r echo = TRUE, warning = FALSE, message=FALSE}
# Plot bar plots of factor variables against churning
bp11 <- plot.barplot(train, train$OnlineSecurity, train$Churn, "Online Security", "Proportion", 15)
bp12 <- plot.barplot(train, train$OnlineBackup, train$Churn, "Online Backup", "Proportion", 15)
bp13 <- plot.barplot(train, train$DeviceProtection, train$Churn, "Device Protection", "Proportion", 15)
bp14 <- plot.barplot(train, train$TechSupport, train$Churn, "Tech Support", "Proportion", 15)
bp15 <- plot.barplot(train, train$StreamingTV, train$Churn, "Streaming TV", "Proportion", 15)
bp16 <- plot.barplot(train, train$StreamingMovies, train$Churn, "Streaming Movies", "Proportion", 15)
grid.arrange(bp11, bp12, bp13, bp14, bp15, bp16, ncol=3)
```

When it comes to services that depend on having **internet service**, there appears to be a significant difference in the proportion of customers who churned for the variables **online security**, **online backup**, **device protection**, and **tech support**, with customers who lack these services churning more often. There doesn't appear to be a significant difference for **streaming TV** or **streaming movies**.

## 4. Other Interesting Relationships

### 4.1 Senior Citizenship, Dependents, and Partner

The variables **senior citizenship**, **dependents**, and **partner** were found to be highly correlated with churning. Since these characteristics are closely related, I wanted to visualize their relationship to see if they are correlated with each other. 

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Plot the relationship between paperless billing and payment method
bp1 <- ggplot(train, aes(y = Partner)) + 
  geom_bar(aes(fill = SeniorCitizen), position = "fill") +
  scale_fill_brewer(palette = "Set1")
bp2 <- ggplot(train, aes(y = Partner)) + 
  geom_bar(aes(fill = Dependents), position = "fill") +
  scale_fill_brewer(palette = "Set1")
bp3 <- ggplot(train, aes(y = SeniorCitizen)) + 
  geom_bar(aes(fill = Dependents), position = "fill") +
  scale_fill_brewer(palette = "Set1")
grid.arrange(bp1, bp2, bp3, ncol=1)
```

**Findings**:

- The proportion of customers who are senior citizens is the same whether or not they have a partner, which means that **senior citizenship** and **partner** are not highly correlated
- Customers who have partners significantly have more dependents than those who do not, which means that **partner** and **dependents** are correlated.
- Customers who are not senior citizens tend to have more dependents than those who are, which means that **senior citizenship** and **dependents** are correlated.

Since **dependents** is correlated with both variables, I will be including it in the model.

### 4.2 Tenure and Contract

Tenure is dependent on contract since customers are probably unable to leave the company without paying significant fees. Thus, those who choose a two-year contract tend to have a higher tenure than those with a month-to-month contract.

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="75%"}
# Plot the relationship between contract and tenure
plot.boxplot(train, train$Contract, train$tenure, "Contract", "tenure")
```

Since these two variables are correlated with each other, I will only be including one of them in the model. To reduce the presence of noise, I decided to go with Contract since it is basically discrete version of tenure. I will also recode it to **Monthly Contract** since the majority of churning customers have *month-to-month* contracts. 

### 4.3 Paperless Billing and Payment Method

Since **paperless billing** and **payment method** are closely related to each other, I decided to plot their relationship. I found that many customers who have paperless billing use electronic checks, which could be the reason why there is a significant difference in the proportion of customers who churned for **paperless billing**. 

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="75%"}
# Plot the relationship between paperless billing and payment method
ggplot(train, aes(x = PaperlessBilling)) + 
  geom_bar(aes(fill = PaymentMethod), position = "dodge") +
  theme(axis.text=element_text(size=16), axis.title=element_text(size=16))
```

Since these two variables are correlated with each other, I will only be including **payment method** in the model. Since the proportion of customers churning appear to be the same for all methods except for electronic check, I will recode payment method to a new variable, **electronic check**, that indicates whether or not a customer used an electronic check.

### 4.4 Monthly Charges and Internet Service

Customers who churn tend to have a higher monthly charge. They also have *Fiber Optic* as their internet service. I looked at the relationship between **internet service** and **monthly charges** and found that customers who have *Fiber Optic* have a higher monthly charge than those who have *DSL*.

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="75%"}
# Plot the relationship between internet service and monthly charges
plot.boxplot(train, train$InternetService, train$MonthlyCharges, "Internet Service", "Monthly Charges")
```


### 4.5 Fiber Optic and DSL

Finally, I wanted to look at difference between the customers who have each of the internet service.

First, I looked at the relationship between the provided services and churning. Customers who have *Fiber Optic* tend to churn more than those with *DSL*. However customers who don't have **online security**, **online backup**, **device protection**, and **tech support** tend to churn more regardless of their internet service.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Subset the data
train.is <- subset(train, InternetService != "No")

# Plot bar plots of services against churning
bp1 <- plot.barplot(train.is, train.is$OnlineSecurity, train.is$Churn, "Online Security", "Proportion", 0) + facet_grid(. ~ InternetService)
bp2 <- plot.barplot(train.is, train.is$OnlineBackup, train.is$Churn, "Online Backup", "Proportion", 0) + facet_grid(. ~ InternetService)
bp3 <- plot.barplot(train.is, train.is$DeviceProtection, train.is$Churn, "Device Protection", "Proportion", 0) + facet_grid(. ~ InternetService)
bp4 <- plot.barplot(train.is, train.is$TechSupport, train.is$Churn, "Tech Support", "Proportion", 0) + facet_grid(. ~ InternetService)
bp5 <- plot.barplot(train.is, train.is$StreamingTV, train.is$Churn, "Streaming TV", "Proportion", 0) + facet_grid(. ~ InternetService)
bp6 <- plot.barplot(train.is, train.is$StreamingMovies, train.is$Churn, "Streaming Movies", "Proportion", 0) + facet_grid(. ~ InternetService)
grid.arrange(bp1, bp2, bp3, bp4, bp5, bp6, ncol=2)
```


Next, I looked at the relationship between **monthly charges** and the provided services. Customers who don't have these services tend to have a lower monthly charge than those who do. However, the monthly charge of customers with *Fiber Optic* are higher than those with *DSL* regardless of the services they have.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Plot box plots of services against monthly charges
bp2 <- plot.boxplot(train.is, train.is$OnlineSecurity, train.is$MonthlyCharges, "Online Security", "Monthly Charges") + scale_fill_brewer(palette = "Set2") + facet_grid(. ~ InternetService)
bp3 <- plot.boxplot(train.is, train.is$OnlineBackup, train.is$MonthlyCharges, "Online Backup", "Monthly Charges") + scale_fill_brewer(palette = "Set2") + facet_grid(. ~ InternetService)
bp4 <- plot.boxplot(train.is, train.is$DeviceProtection, train.is$MonthlyCharges, "Device Protection", "Monthly Charges") + scale_fill_brewer(palette = "Set2") + facet_grid(. ~ InternetService)
bp5 <- plot.boxplot(train.is, train.is$TechSupport, train.is$MonthlyCharges, "Tech Support", "Monthly Charges") + scale_fill_brewer(palette = "Set2") + facet_grid(. ~ InternetService)
bp6 <- plot.boxplot(train.is, train.is$StreamingTV, train.is$MonthlyCharges, "Streaming TV", "Monthly Charges") + scale_fill_brewer(palette = "Set2") + facet_grid(. ~ InternetService)
bp7 <- plot.boxplot(train.is, train.is$StreamingMovies, train.is$MonthlyCharges, "Streaming Movies", "Monthly Charges") + scale_fill_brewer(palette = "Set2") + facet_grid(. ~ InternetService)
grid.arrange(bp2, bp3, bp4, bp5, bp6, bp7, ncol=3)
```

Based on these observations, I can conclude that the reason why customers with *Fiber Optic* tend to churn more isn't due to the quality of their services, but because of the higher monthly charge. Also, customers who have these services tend to stay more in the company despite paying more for these services. 

## 5. Conclusion

This section focused on exploring and analyzing the data to determine which variables will be suitable for predictive modeling. I wanted to avoid running t-tests and chi-squared tests due to the multiple comparisons problem, where the occurrence of finding a significant result by random chance increases. Thus, I focused on using visualizations and numerical summaries to determine which predictors to include in the model. I also wanted to avoid issues with collinearity, so I made sure to explore the relationships between the predictors.

Based on these results, I decided to include the following variables in the model:

1. **Monthly Charges**
2. **Dependents**

I will also include two new variables recoded from existing variables:

1. **Monthly Contract** - whether or not a customer has a month-to-month contract.
2. **Electronic Check** - whether or not a customer's payment method is Electronic check.
3. **Has Service** - whether or not a customer as at least one of the following: online security, online backup, device protection, and tech support. 

